---
title: "Using the tidyverse with Databases - Part II"
description: |
  I was serious, you don't need to be a SQL expert to work with Databases through R
author:
  - name: Vebash Naidoo
    url: https://twitter.com/Sciencificity
    affiliation: R-Ladies Johannesburg
    affiliation_url: https://twitter.com/RLadiesJozi
date: 12-20-2020
output:
  distill::distill_article:
    toc: true
    toc_depth: 3
    self_contained: false
categories:
  - SQL
  - databases
  - tidyverse
  - dbplyr
preview: Paper.R_Programming.4.png
---

<img src = Paper.R_journal.33_.png width = 600px>

```{css, echo=FALSE}
.main-container {
    max-width: 600px !important;
}

pre {
  max-height: 800px !important;
  max-width: 600px !important;
  overflow-y: auto !important;
  overflow-x: scroll !important;
}

pre[class] {
  max-height: 100px;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
library(flair)
```

```{r allow-code-amend-size, echo=FALSE}
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```

```{r, include=FALSE}
options(width = 160)
local({
  hook_output <- knitr::knit_hooks$get('output')
  knitr::knit_hooks$set(output = function(x, options) {
    if (!is.null(options$max.height)) options$attr.output <- c(
      options$attr.output,
      sprintf('style="max-height: %s;"', options$max.height)
    )
    hook_output(x, options)
  })
})
```


# Part I

In [Part I](https://sciencificity-blog.netlify.app/posts/2020-12-12-using-the-tidyverse-with-databases/) we made a connection to a SQLite DB.

# What are we tackling in Part II?

- We'll do more advanced `dplyr` `r emoji::emoji("package")` workflows.
- We'll **collect** the data from the query.

<aside>
`collect`ing data is when we bring the data into R. If you recall in the Part I post, all computation was happening on the DB itself, and not pulled into R.
</aside>

The [project on GitHub](https://github.com/sciencificity/intro-to-dbs-r), has the example SQLite database, the slides, and some code files.


# Connect, and remind ourselves what we're working with

## Make a connection

As always, our __first__ step is to <span style="color: #008080;background-color:#e4a199"><strong>connect to the database</strong></span>.

```{r}
library(DBI) # main DB interface
library(dplyr) 
library(dbplyr) # dplyr back-end for DBs

con <- dbConnect(drv = RSQLite::SQLite(), # give me a SQLite connection
        dbname = "data/great_brit_bakeoff.db") # To what? The DB named great_brit_bakeoff.db

dbListTables(con) # List me the tables at the connection
```

## Let's get familiar with our data

In the dataset we have:

- <span style="color: #5196b4;background-color:#f4e4e7">results<a href="https://bakeoff.netlify.app/reference/results.html"></a></span> which tells us [how each baker did](https://bakeoff.netlify.app/reference/results.html) in the episode - `IN`, `OUT`, `WINNER` etc.
- <span style="color: #5196b4;background-color:#f4e4e7">baker_results<a href="https://bakeoff.netlify.app/reference/baker_results.html"></a></span> which tells us [a bit more](https://bakeoff.netlify.app/reference/baker_results.html) about the baker's performance over the entire series - how many times did they win `STAR BAKER`, how many times they placed in the top 3 etc.

Let's get an idea of what is in each table. 

### results

```{r res}
tbl(con, "results") %>% # Reach into my connection, and "talk" to results table
  head(10) %>%          # get me a subset of the data
  # sometimes if there are many columns, some columns are hidden, 
  # this option prints all columns for us
  print(width = Inf)    
```

```{r}
tbl(con, "results") %>% 
  count(result, sort = TRUE) # what categories of "result" are there?
```

### baker_results

```{r}
tbl(con, "baker_results") %>% # Reach in and "talk" to baker_results
  head() %>%                  # get a glimpse of data
  collect() %>%               # bring that glimpsed data into R 
  DT::datatable(options = list(scrollX = TRUE)) # force DT horizontal scrollbar
```
Notice the use of the <span style="color: #008080;background-color:#e4a199">`collect()`</span> function in the code above. I wanted us to be able to get a full glimpse of the data in a nice table, and hence I brought the first few rows of data into R by using the <span style="color: #008080;background-color:#e4a199">`collect()`</span> function. This allowed me to then use `datatable` to display the results a bit better, than the `print(width = Inf)` alternative.

# What are we interested in?

Let's say we want to see how the __`WINNER`__ and __`RUNNER-UP`(s)__ did in the series they appeared in.

To do that we need to get all the <span style="color: #5196b4;background-color:#f4e4e7">baker_results</span> for the `WINNER` and `RUNNER-UP`.

## Joining data

When doing joins we want to find the common columns across the two tables that we can `join` on.

- In <span style="color: #5196b4;background-color:#f4e4e7">results<a href="https://bakeoff.netlify.app/reference/results.html"></a></span> we have `series` and `baker`.
- In <span style="color: #5196b4;background-color:#f4e4e7">baker_results<a href="https://bakeoff.netlify.app/reference/baker_results.html"></a></span> we have `series` and `baker`.
- `r emoji::emoji("question")` `r emoji::emoji("thinking")` Why didn't I also choose the `episode` column of <span style="color: #5196b4;background-color:#f4e4e7">results</span> as a join column? <br> `r emoji::emoji("white_check_mark")` Yes, that column is not in `baker_results`, since `baker_results` contains data about how the baker did overall in the series they appeared, that is, one row per baker. <br>The `results` data however, contains info per baker, per episode, for the series they appeared in - i.e. if they flopped (were `OUT` `r emoji::emoji("wink")`), in the second episode of a series that contained 10 episodes, their name would appear 10 times in the `results` table, but their "result" value will be `NA` from episode 3 onwards.

### Remember the `tbl(con, "tbl_name")` always

I'd like to bring to your attention the use of `tbl(con, "table_1")` and `tbl(con, "table_2")` in the __join__ function. 

We must always keep this in mind, because <span style="color: #5196b4;background-color:#f4e4e7">baker_results</span> and <span style="color: #5196b4;background-color:#f4e4e7">results</span> don't exist in R yet. We're talking to those tables in our relational database management system (RDBMS), so we __always__ have to do so through our __connection__.

```{r}
set.seed(42)
tbl(con, "baker_results") %>% # use connection to "talk" to baker_results
  inner_join(tbl(con, "results"), # use connection to "talk" to results and join both tables 
        by = c('baker' = 'baker',
               'series' = 'series')) %>% # join criteria 
  collect() %>% # get it into R
  sample_n(size = 3) %>% # take a random sample
  print(width = Inf) # print all columns
```

Notice that all columns of <span style="color: #5196b4;background-color:#f4e4e7">baker_results</span> appear first and then we have the "extra" columns from <span style="color: #5196b4;background-color:#f4e4e7">results</span> i.e. `episode` and `result`.

#### Common mistake

I included the above to show that each time we "talk" to a table we must do so through our connection, because I often make the mistake of not including the `tbl(con, "name_of_tbl_i_am_joining")` in the __join__ function. I, more times than I care to admit `r emoji::emoji("woman_facepalming")`, __incorrectly__ write:

<pre>
  tbl(con, "table1") %>% # correct -> "talk" through connection
     left_join(table2,   # incorrect -> forgot to use the connection
          by = c("tbl1_name1" = "tbl2_name1"))
</pre>

I would like to help you, not repeat my mistake `r emoji::emoji("confused")`, so heads up <span style="color: #FFE4E1;background-color:#FD5800">AVOID THE FOLLOWING</span> `r emoji::emoji("stop_sign")`:

```{r, error = TRUE}
tbl(con, "baker_results") %>% # use connection to "talk" to baker_results
  inner_join(results,  # OOPS! I forgot the tbl(con, "results")
        by = c('baker' = 'baker',
               'series' = 'series'))  
```


## Collect

Ok, let us now do our entire pipeline, and only bring the data into R when we've got what we're looking for.

We need to:

- Join the tables
- Filter the data for `WINNER` and `RUNNER-UP` in the __result__ column.
- Select only the columns we're interested in.

```{r}
(final_query <- tbl(con, "baker_results") %>% # use connection to "talk" to baker_results
  inner_join(tbl(con, "results"), # use connection to "talk" to results and join both tables 
        by = c('baker' = 'baker',
               'series' = 'series')) %>% # join criteria 
  filter(result %in% c("WINNER", "RUNNER-UP")) %>% # filter rows we're interested in
  select(series, baker:percent_technical_top3,
         result))
```

The above code just sets up the query that will be executed should we run (Ctrl + Enter) `final_query` in R (hence the __lazy query [?? x 24]__ in the output). No data is collected (i.e. present in your local R environment) as yet.

### What does the query look like?

```{css, echo=FALSE}
.scroll-100 {
  white-space:pre !important;
  max-width: 100px;
  overflow-x: scroll;
  background-color: inherit;
}
code.r{
  font-size: 10px;
}
```


```{r, size = "tiny", class.output="scroll-100"}
final_query %>% 
  show_query() 
```

\normalsize


### Bring it into R

Now finally, we are ready to bring our filtered and joined data into R by using `collect()`.

```{r}
(top_performers <- final_query %>% 
  collect())
```

How about that? Notice the __A tibble: 24 x 24__! R now, has the data in it's local environment, and can definitively tell us it knows there are 24 observations (no more _lazy query_) `r emoji::emoji("smile")`.

### Visualise Data

Now that we have finalised what data we wanted from our RDBMS, executed our query, and collected the data into our R environment we can do further processing, create plots for reports etc.

I am interested in understanding how did the winner and runner-up(s) of series 6 do across the season in terms of technical challenges etc.?

```{r}
library(tidyverse)

top_performers %>% 
  # filter for season we're interested in
  filter(series == 6) %>%
  # format baker nicely so we see winner, then runner-up(s)
  mutate(baker_name = factor(str_glue("{result} - {baker}")),
         baker_name = fct_rev(baker_name)) %>% 
  # let's convert all the tech info cols to be a metric name, 
  # and put the value in the value column 
  # (by default values_to = "value" in pivot_longer())
  pivot_longer(cols = c(star_baker:technical_median),
               names_to = "metric") %>% 
  mutate(metric = fct_reorder(metric, value)) %>% 
  ggplot(aes(x = value, y = metric)) +
  geom_col(fill = "#727d97") +
  facet_wrap(~ baker_name) +
  labs(title = str_glue("Metrics for Season ",
    "{top_performers %>%  filter(series == 6) %>%
    select(series) %>% distinct()}'s Winner and Runner-Up(s)"),
    y = "") +
  theme_light() +
  theme(panel.spacing = unit(1, "lines")) +
  theme(strip.background =element_rect(fill="#f4e4e7"))+
  theme(strip.text = element_text(colour = "#5196b4"))
```

Given that _Nadiya_ was a technical winner more times than the other contestants, and that her `technical_lowest` was better (higher number is better) it looks like she had a good run throughout the series, and was a deserved winner.

## Done? Remember to disconnect!

Good housekeeping means always remembering to disconnect once you're done.

```{r}
dbDisconnect(con) # closes our DB connection
```

# Still to come

- Setting up an external DB in PostgreSQL / MySQL.
- Connecting to said DB, and running SQL workloads on it.

## Acknowledgements

- `The Great British Bake Off` data from [Dr. Alison Hill](https://bakeoff.netlify.app/reference/index.html).
- [Edgar Ruiz's](https://rstudio.com/speakers/edgar-ruiz/) database work, and teachings.








