---
title: "Using the tidyverse with Databases - Part I"
description: |
  You don't need to be a SQL expert to work with Databases through R
author:
  - name: Vebash Naidoo 
    url: https://twitter.com/Sciencificity
    affiliation: R-Ladies Johannesburg
    affiliation_url: https://twitter.com/RLadiesJozi
date: 2020-12-12
output:
  distill::distill_article:
    toc: true
    toc_depth: 3
    self_contained: false
categories:
  - SQL
  - databases
  - tidyverse
  - dbplyr
preview: data2-unsplash.jpg
---

<img src = data.jpg width = 600px>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      warning = FALSE)
library(flair)
```

# Background
On the 1^st^ of December I certified as an [RStudio Tidyverse Instructor](https://bit.ly/my-rstudio-cert-profile). As part of the process I prepared a sample lesson that was delivered during the exam. In a series of posts I will go through an extension of that sample lesson.

# Intro to using Databases in R, with Tidyverse tools (Part I)

## Motivation

You often have data in a corporate or institutional database (often termed relational database management system, or `RDBMS`). The language to talk with such a database (DB) is some variant of the language SQL (Structured Query Language). I say some variant, since a vendor providing an RDBMS may adopt the standardised SQL, but they may also extend it with their own variations.

If you use different RDBMS's from different vendors you need to understand how to query that DB using the particular SQL dialect used by that vendor. For example, we use PostgreSQL, and we also use PrestoDB (for Athena on AWS). `JSON` fields in the PostgreSQL DB are ingested as structured nested arrays in AWS and hence when querying the same field we use different query syntax.

You know R, especially the `dplyr` `r emoji::emoji("package")`. Even though the `dplyr` `r emoji::emoji("package")` is so well written to mimic the SQL syntax - __select()__, __group_by()__, __left_join()__ etc. there is still a cognitive load when you switch between using R syntax, and SQL syntax (ask me, who has often written `==` in SQL syntax on Athena only to wonder why I am getting an error `r emoji::emoji("zipper_mouth_face")`).

You only have so much memory in your local environment, and may want your RDBMS to do the heavy lifting (most of the computation), and only pull data into R when you need to (e.g. pull in aggregated data to create plots for a report).

In this tutorial you will learn how to use `dbplyr`, which is a database back-end of `dplyr`, to execute queries directly in your RDBMS all the while writing R tidyverse syntax `r emoji::emoji("open_mouth")` `r emoji::emoji("star2")`.

## Connecting to a Database

### Packages needed

- DBI: This is the main package which is a Database Interface for R.<br> `install.packages("DBI")`
- dbplyr: The `r emoji::emoji("package")` we may use to talk to our database, all the while using `dplyr` syntax in our code.<br>`install.packages("dbplyr")`
- dplyr: To `select()`, `group_by()` as well as "talk", via the connection we will establish, to a database table.
- RSQLite: DBI compliant package that allows a connection to a lightweight single user SQLite DB.<br>`install.packages("RSQLite")`
- Other DBI compliant R packages you may need for your use case - e.g. `RPostgres` is for connecting to a PostgreSQL RDBMS. *Note: Not used in this tutorial (brief overview in slides), and [RStudio's DB Website](https://db.rstudio.com/) is a treasure trove of information* `r emoji::emoji("woman_technologist")`.
- odbc: This is a DBI compliant interface to allow you to interface to your RDBMS using an odbc driver.*Note: Not used in this tutorial (brief overview in slides), and again [RStudio's DB Website](https://db.rstudio.com/) has useful information in this regard too*!<br>`install.packages("odbc")`

### The database we'll use

[Alison Hill](https://alison.rbind.io/) created a dataset for [_The Great British Bake Off_](https://github.com/apreshill/bakeoff). I used the data from her package to create a SQLite DB that we will use to practise on.

### Connecting to a SQLite DB

For the purposes of this tutorial we will connect to a SQLite DB that I created. To talk to the DB we need to first __make a connection__.

1. Load the DBI package:
    ```
    library(DBI)
    ```
2. Make a connection:
    ```
    con <- dbConnect(RSQLite::SQLite(), "mydb.db")
    ```

```{r}
library(DBI) # main DB interface
library(dplyr) 
library(dbplyr) # dplyr back-end for DBs

con <- dbConnect(drv = RSQLite::SQLite(), # give me a SQLite connection
        dbname = "data/great_brit_bakeoff.db") # To what? The DB named great_brit_bakeoff.db
summary(con) # What do we have? 
```

Ok, we have successfully managed to connect to the DB.

## Take a look around

Alright, we've made a connection now what? Let's have a look around using some DBI functions. 

### DBI Functions

```{r}
dbListTables(con) # List me the tables at the connection
dbListFields(con, # Reach into my connection and ...
      "bakers")   # tell me what fields does the `bakers` table have?
res <- dbSendQuery(con, "SELECT * FROM bakers LIMIT 3") # Execute a query
dbFetch(res) # get the result
dbClearResult(res) # clear the result
```

### `dplyr` Functions

Ok, let's flex our `dplyr` skills `r emoji::emoji("smile")`.

- `tbl(con, "name_of_table")` says using my connection "con" go execute a `SELECT * FROM name_of_table`.

```{r}
tbl(con, "bakers") # let's have a look - works like glimpse()
tbl(con, "bakers") %>% 
  head(3) # "SELECT * FROM bakers LIMIT 3"
```

Notice that each time we used <span style="color: #008080;background-color:#9FDDBA">`con`</span>. We use our connection __con__ to "talk" to our database, and we will use it throughout, even in our _tidy_ pipelines.

In `baker_results` we have the baker with their details, as well as their standing in the series competition. Let's say we wanted to know if the winners came from different areas in Britain, or from some areas in particular.

```{r}
dbListFields(con, "baker_results")
tbl(con, "baker_results") %>% 
  select(series, baker, hometown, series_winner) # normal dplyr select
```

Notice how our `dplyr` __select()__ has trimmed down the number of columns from 24 to 4.

Time to see if our winning bakers hail from similar hometowns or not.

```{r}
tbl(con, "baker_results") %>% 
  select(series, baker, hometown, series_winner) %>% 
  filter(series_winner == 1) %>% # normal dplyr filter
  count(hometown, sort = TRUE)   # normal dplyr count
```

Looks like our winners are from very different areas.

### Have a look at the SQL behind the scenes

Our `dplyr` syntax is converted behind the scenes to SQL which is executed directly on the DB table.

```{r show_query}
tbl(con, "baker_results") %>% 
  select(series, baker, hometown, series_winner) %>% 
  filter(series_winner == 1) %>% 
  count(hometown, sort = TRUE) %>% 
  show_query()
```

That's pretty nifty, and helps you learn SQL in the process - it is always good to know, and read SQL, which is used extensively in work and institutional environments.

### What's with this `lazy query` / `??` we're seeing?

You might have noticed that our pipelines have a:

 - <span style="color: #008080;background-color:#9FDDBA">Source:   table<xxx> [?? x 5]</span> OR
 - <span style="color: #008080;background-color:#9FDDBA">Source:   lazy query [?? x 4]</span>

in the output results.

#### Why is this?

- Behind the scenes the `dplyr` code we write, gets converted to __SQL__ (the particular dialect of the DB we're connecting to).
- The SQL is executed directly on the DB. I.e. the data is not first brought into R and manipulated in R. Instead we send the query to the DB, and the DB does the computation. 
- When we use a `dplyr` pipeline to execute a query on the DB, the DB does the computation and sends us back a glimpse of the executed end result. 
- It is for this reason, we see the __`??`__ - we're being told "Hey, I executed this query (your pipeline converted to SQL) on the DB at connection `con`, and here's a snippet of the output, but I did not get the meta-info on how many rows were produced, I just know that I got x_number of columns in the output."
- We will see in the next part of the tutorial how to get the info into R, and understand how many rows were produced etc.

## Done? Remember to disconnect!

Good housekeeping means always remembering to disconnect once you're done.

```{r}
dbDisconnect(con) # closes our DB connection
```

To see that is indeed the case let's see if we can use the connection further ...

```{r, error = TRUE}
dbListTables(con)
```


# Slides

If you'd like to dive in please check out the slides, and the [project on GitHub](https://github.com/sciencificity/intro-to-dbs-r).

```{r share-again, echo=FALSE}
xaringanExtra::use_share_again()
```

```{r embed-xaringan, echo=FALSE, fig.cap="[Slides](slides/index.html)"}
xaringanExtra::embed_xaringan(url = "slides/index.html", ratio = "16:9")
```

# Still to come

- More `dplyr`, for example joining tables.
- Getting the data into R.

## Acknowledgements

- Slide inspiration, as well as some of the background pictures are from [Dr. Alison Hill](https://alison.rbind.io/).
- `The Great British Bake Off` data from [Dr. Alison Hill](https://alison.rbind.io/).
- [Edgar Ruiz's](https://rstudio.com/speakers/edgar-ruiz/) database work, and teachings.
